Regularising Lotka--Volterra interaction matrices by species grouping: optimal groupings are unpredictable

### 2022-02-21 article.rnw
adding one more by-hand calculation about null co-classification
adding a note regarding an idea for another figure
adding some comments to the discussion
Add Daniel as an author.
In doing so, remove him from the acknowledgements.
Fix a mangled sentence break.

%Title Page
\title{Regularising Lotka--Volterra interaction matrices by species grouping: optimal groupings are unpredictable}
\author{Christopher R. P. Brown}
\affil{Centre for Integrative Ecology, School of Biological Sciences, University of Canterbury, Christchurch, New Zealand}
\date{}

% DBS: We should have a conversation regarding authorship. To me, it makes perfect sense thus far that you are the sole author. Depending on how revisions go and what anyone else or I do or do not contribute, it may be appropriate for this to change. In many such instances, I often think that the lead author (i.e., you) will have the best feeling for this relative to the criteria set out in the Vancouver guidelines for authorship.

@@

The Lotka--Volterra model is not linear in the parameters.
With the rows of the interaction matrix ungrouped, the regression could be reparameterised to be linear in the parameters by treating $r_i \alpha_{ij}$ as a single regression parameter for each $i,j$ pair.
With the rows grouped, however, this becomes impossible since each species still has an independent growth rate.
As such, the models were fitted by the Gauss--Newton method for non-linear least squares.
%The models were fitted to the data using maximum likelihood regression techniques.
%Fitting it using precisely a Lotka--Volterra model prevented the use of a link function, rendering most generalised regression techniques infeasible.
I assumed a normal error distribution of the response variable. and 
to prevent heteroskedasticity and improve normality, I transformed total fecundity to \emph{per capita} fecundity, which corresponds to the following model:
\[
	\frac{N_{i,t+1}}{N_{i,t}} = r_i \left(1 - \sum_j \alpha_{ij} N_{j,t}\right)
.\]
% DBS: CRPB to alter equation above consist with changes to previous equations.

@@

In all, no pair of species can confidently be said to be coclassified, in the row or column groupings; no probability of coclassification exceeded \SI{60}{\percent}.
Some pairs of species were far more likely to be coclassified than others, however, suggesting some structure to the best groupings.
% DBS: As I have noted before, I'd like to know what the null expectation is for this before claiming confidence of being co-classified or not being co-classified. Firstly, I think that the random expectation is much lower than one might otherwise suspect. (For three species, it is already 2/5.) Second, the conventional "ecological" null is actually that the probability is 0, in which case it doesn't take much co-classification probability to see the glass as half full and argue that grouping is a much more parsimonious approach than the contrary.

@@

@@

The results show that grouping the interaction coefficients of species was likely to give a more parsimonious model (probability \probnum{\Sexpr{sum(brute_data$statistics$aic_weight[brute_data$statistics$aic < brute_data$fully_separated_statistics$aic])}}) than leaving them ungrouped, but not confidently so.
Which grouping might give the most parsimonious model also could not be determined with any confidence.
The intransitivity of coclassification in some instances even suggests that there may not be a single ``best'' grouping, but rather two or more quite distinct and mutually exclusive groupings of similar validity.

This stands in contrast to the popular wisdom in ecology that species should form some sort of clear groups, be they the functional groups so often used \parencite{blondel2003guilds}, or taxonomic groups per a relationship between phylogeny and ecological niche \parencite{godoy2014phylogenetic}.
It also contrasts with previous studies that have succeeded in finding clear groups, such as \textcite{allesina2009food}, who used AIC to find the best grouping of species, but in a binary food web rather than a qualitative competitive context.
Perhaps species simply do not group to the same extent in a competitive context.
The results suggest that this is unlikely, as grouping was still likely to increase parsimony.
Further, there is no reason to suspect that species would not have clear similarities or dissimilarities in their ecological niches, and therefore their competitive effects and responses.
Perhaps the lack of clarity in the groupings is instead a product of the study system chosen.

I chose a study system of \emph{Drosophila} partially on the basis that it would allow clear analysis of the species' phylogenetic relationships.
Phylogeny proved a poor predictor of optimal grouping, as previous studies have suggested that it might \parencite{godoy2014phylogenetic,cahill2008does}.
A study system with species spanning multiple functional groups, however, might have shown clearer groupings.
While functional groups have previously shown to be poor predictors of competitive interactions \parencite{bret2008plant}, studies in other contexts have succeeded in isolating functional groups by selecting for model parsimony \parencite{allesina2009food}.
In further support of functional groups over phylogeny, \textcite{uriarte2004spatially} found that grouping based on light guilds (akin to functional groups; \cite{blondel2003guilds}) produced a parsimonious model more frequently than grouping taxonomically (with both groupings defined \emph{a priori}).
Perhaps an attempt to produce functional groups based upon competitive interactions is warranted, applying the framework I present here to a set of species spanning multiple functional groups.

Alternatively, it should be noted that the success of \textcite{allesina2009food} in isolating groups using AIC occurred in much larger study systems, with 25 or more species.
Alas, the brute force method I employed in this study has a very poor algorithmic complexity, preventing the application of the method to studies with more than seven or eight species.
As such, a better method would be required in order to test for grouping of species interactions among a larger set of species.

In light of the failure to produce clear groups, perhaps the most curious aspect of the results is the disagreement between the likely coclassifications of species in the groupings of rows and of columns.
This is perhaps not surprising in hindsight, for there is no necessity that species respond to competition in the same manner that they cause it, especially in competition for certain resources such as space (the limiting resource presumed in these data) or light.
Yet the assumption that species should group equivalently in these two regards is frequently made, often implicitly and without comment \parencite[e.g.][]{martorell2014testing,ives2003estimating,gross2015stability}.
My results recommend more explicit questioning, and even testing, of that assumption when building future species interaction models.
%TODO: Discuss the fact that species were coclassified more in the rows than in the columns.

@@

\section*{Acknowledgements}

I thank Daniel B. Stouffer for advice throughout the study, and particularly for making me aware of Akaike weights and suggesting the use of coclassification matrices.

>>>>

%Title Page
\title{Regularising Lotka--Volterra interaction matrices by species grouping: optimal groupings are unpredictable}
\author{Christopher R.\ P.\ Brown}
\author{Daniel B.\ Stouffer}
\affil{Centre for Integrative Ecology, School of Biological Sciences, University of Canterbury, Christchurch, New Zealand}
\date{}

@@

The Lotka--Volterra model is not linear in the parameters.
With the rows of the interaction matrix ungrouped, the regression could be reparameterised to be linear in the parameters by treating $r_i \alpha_{ij}$ as a single regression parameter for each $i,j$ pair.
With the rows grouped, however, this becomes impossible since each species still has an independent growth rate.
As such, the models were fitted by the Gauss--Newton method for non-linear least squares.
%The models were fitted to the data using maximum likelihood regression techniques.
%Fitting it using precisely a Lotka--Volterra model prevented the use of a link function, rendering most generalised regression techniques infeasible.
I assumed a normal error distribution of the response variable and to prevent heteroskedasticity and improve normality, I transformed total fecundity to \emph{per capita} fecundity, which corresponds to the following model:
\[
	\frac{N_{i,t+1}}{N_{i,t}} = r_i \left(1 - \sum_j \alpha_{ij} N_{j,t}\right)
.\]
% DBS: CRPB to alter equation above consist with changes to previous equations.

@@

In all, no pair of species can confidently be said to be coclassified, in the row or column groupings; no probability of coclassification exceeded \SI{60}{\percent}.
Some pairs of species were far more likely to be coclassified than others, however, suggesting some structure to the best groupings.
% DBS: As I have noted before, I'd like to know what the null expectation is for this before claiming confidence of being co-classified or not being co-classified. Firstly, I think that the random expectation is much lower than one might otherwise suspect. (For three species, it is already 2/5; for four species, it is 5/15.) Second, the conventional "ecological" null is actually that the probability is 0, in which case it doesn't take much co-classification probability to see the glass as half full and argue that grouping is a much more parsimonious approach than the contrary.

@@

% DBS: I think a good figure would be to show the ungrouped estimated alpha matrix in one panel and the alpha matrix averaged across all partitions in a second panel, with lines drawn in the second to reflect the "optimal" partitioning of the co-classification matrices.

@@

% DBS: I would start this paragraph with a restatement of the overall premise.
The results show that grouping the interaction coefficients of species was likely to give a more parsimonious model (probability \probnum{\Sexpr{sum(brute_data$statistics$aic_weight[brute_data$statistics$aic < brute_data$fully_separated_statistics$aic])}}) than leaving them ungrouped, but not confidently so.
Which grouping might give the most parsimonious model also could not be determined with any confidence.
% DBS: The two preceding sentences are very glass half empty. I'm also not sure that the results are really so equivocal. What makes you feel this way?
The intransitivity of coclassification in some instances even suggests that there may not be a single ``best'' grouping, but rather two or more quite distinct and mutually exclusive groupings of similar validity.
% DBS: In my mind, I would steal the preceding sentence and spin it potentially spin it into its own paragraph later since it speaks strongly to a mixed-membership approach which could be sold as a "future direction".

% DBS: Avoid imprecise "this" or "that" as subjects. Always say "This result" or something similar. Also, avoid starting a new paragraph that so strongly forms a continuation of the previous one.
This stands in contrast to the popular wisdom in ecology that species should form some sort of clear groups, be they the functional groups so often used \parencite{blondel2003guilds}, or taxonomic groups per a relationship between phylogeny and ecological niche \parencite{godoy2014phylogenetic}.
It also contrasts with previous studies that have succeeded in finding clear groups, such as \textcite{allesina2009food}, who used AIC to find the best grouping of species, but in a binary food web rather than a qualitative competitive context.
Perhaps species simply do not group to the same extent in a competitive context.
The results suggest that this is unlikely, as grouping was still likely to increase parsimony.
Further, there is no reason to suspect that species would not have clear similarities or dissimilarities in their ecological niches, and therefore their competitive effects and responses.
Perhaps the lack of clarity in the groupings is instead a product of the study system chosen.

I chose a study system of \emph{Drosophila} partially on the basis that it would allow clear analysis of the species' phylogenetic relationships.
Phylogeny proved a poor predictor of optimal grouping, as previous studies have suggested that it might \parencite{godoy2014phylogenetic,cahill2008does}.
A study system with species spanning multiple functional groups, however, might have shown clearer groupings.
While functional groups have previously shown to be poor predictors of competitive interactions \parencite{bret2008plant}, studies in other contexts have succeeded in isolating functional groups by selecting for model parsimony \parencite{allesina2009food}.
In further support of functional groups over phylogeny, \textcite{uriarte2004spatially} found that grouping based on light guilds (akin to functional groups; \cite{blondel2003guilds}) produced a parsimonious model more frequently than grouping taxonomically (with both groupings defined \emph{a priori}).
Perhaps an attempt to produce functional groups based upon competitive interactions is warranted, applying the framework I present here to a set of species spanning multiple functional groups.
% DBS: Prolix warning! ;)

Alternatively, it should be noted that the success of \textcite{allesina2009food} in isolating groups using AIC occurred in much larger study systems, with 25 or more species.
% DBS: Personally, I try to make links to previous studies more self contained. Since the Allesina paper and its use of AIC was brought up earlier, it strikes me as repetitive to lean on it alone again here.
Alas, the brute force method I employed in this study has a very poor algorithmic complexity, preventing the application of the method to studies with more than seven or eight species.
As such, a better method would be required in order to test for grouping of species interactions among a larger set of species.

% DBS: The paragraph below is a very good example of text that could easily be simplified for clarity without losing important content.
In light of the failure to produce clear groups, perhaps the most curious aspect of the results is the disagreement between the likely coclassifications of species in the groupings of rows and of columns.
This is perhaps not surprising in hindsight, for there is no necessity that species respond to competition in the same manner that they cause it, especially in competition for certain resources such as space (the limiting resource presumed in these data) or light. % DBS: Some definite refs missing here.
Yet the assumption that species should group equivalently in these two regards is frequently made, often implicitly and without comment \parencite[e.g.][]{martorell2014testing,ives2003estimating,gross2015stability}.
My results recommend more explicit questioning, and even testing, of that assumption when building future species interaction models.
%TODO: Discuss the fact that species were coclassified more in the rows than in the columns.

@@
